Field-informed Reinforcement Learning for Learning Large-Scale Collective Tasks

### 2023-05-15 paper.tex
try to answer to lukas questions

On these minimal operators, 
 it is then possible to build \emph{self-organizing} coordination blocks.
 One of the founding blocks is the \emph{gradient} essential for 
 \emph{information} flows~\cite{DBLP:conf/saso/WolfH07}.
%
This operator generates a numerical field that represents the minimum distance from a source zone. 
%
In other words, it maps a Boolean field (where ``true'' indicates the presence of a source node and ``false'' indicates its absence) to a distance field that indicates the proximity to the nearest source. In ScaFi (a scala implementation of aggregate computing) it is simply expressed as:
\begin{lstlisting}
def gradient(source: Boolean): Double
\end{lstlisting}
Along the gradient, it is possible to cast information within the system, accumulating value during the programming. This pattern is called \emph{gradient-cast} (or \emph{G}) 
\begin{lstlisting}
def G[V](source: Boolean, value: V, acc: V => V): V
\end{lstlisting}
\ga{more description here}
this is used in the use case to accumulate the distance from the phenomena that is tracked
\lukas{still not sure how the coordination works or if it has been explained in the above paragraph.}
\subsection{Graph Neural Networks}
%Over the years, several different neural networks have been proposed to solve specific tasks, ranging from 
%  simple \ac{MLP}~\cite{werbos1975beyond} to \ac{CNN}~\cite{oshea2015introduction}.
%  \lukas{more examples? }
%%
%Whereas \ac{CNN} is designed to be used with image and spatial-like data (e.g., audio), 
 \ac{GNN} is a novel neural network model used to process graph-structured data with deep learning approaches.
%
Let $G = (V, E)$ a graph  
where $E$ defines the neighbourhood relations for each participating node, 
and $V$ identifies the nodes present in the graph. %at that time.  \lukas{no time in the definition}
% 
Each node $v \in V$ is associated with an observation (or feature set) $o_v$.
%\lukas{Over time this graph $G$ can change and therefore we define $G_t$ as the graph $G$ at a specific time $t$.} 
%
Given $o_v$, %a feature vector associated with each node $v \in V$, 
 the goal of a \ac{GNN} is to learn the node embedding $h_v$ for each node $v \in V$. The node embedding $h_v$ describes the node in the network and summarises the geometric properties of the graph in this location, allowing for comparison of various nodes in the graph.
 In modern \acp{GNN}, the node embedding $h_v$ is computed by aggregating information from the node's neighbours $N(v)$,
  and then combining it with the node's current embedding $h_v$.
 The graph is further partitioned into layers with the node $v$ at its bottom layer $0$. With each traversal of an edge $e \in E$ from $v$, the number of layers increased
Formally, a \ac{GNN} can be defined by two phases:
\begin{equation}
a_{v}^{(k)}=\operatorname{AGGREGATE}^{(k)}\left(\left\{h_{u}^{(k-1)}: u \in \mathcal{N}(v)\right\}\right)   
\end{equation}

\begin{equation}
h_{v}^{(k)}=\operatorname{COMBINE}^{(k)}\left(h_{v}^{(k-1)}, a_{v}^{(k)}\right)
\end{equation}
where $h_{v}^{k}$ is the embedding of node $v$ within the $k$-th layer ($h_v^0 = o_v$), 
 $\mathcal{N}(v)$ is the set of neighbours of node $v$ computed from $E$.
%
The differential part comes into play in the $\operatorname{COMBINE}$ function, 
 which is usually a differentiable function such as a neural network layer.
%
$\operatorname{AGGREGATE} $ instead is a function that aggregates the information from the neighbours of a node $v$
  and it could be a simple sum, max or sum of products. More complex aggregation functions are available~\cite{pellegrini2020learning}.
%
This formulation allows GNNs to effectively process and extract features from graph-structured data by iteratively aggregating and transforming information from the node's neighbours.
\lukas{do we need definitions of $o_v$, $h_v^*$ over time $t$?}

@@

This model can be used then to express the evolution of the system in time.
Specifically, starting from a global state $\mathcal{S}^P_t$, the next state $\mathcal{S}^P_{t+1}$ is defined as:
\begin{equation}
\mathcal{A}^P_t = \pi(\xi(S^P_t))
\end{equation}
\begin{equation}
\mathcal{S}^P_{t+1} = \mathcal{T}(\mathcal{S}^P_t, \mathcal{A}^P_t)
\end{equation}
Given a time $t$, the system can be also represented as a graph $G_t = (V_t, E_t)$, 
  where $E_t$ is built from $\mathcal{N}$.
  Each node is then decorated with the local observation $o_v$. 
  Therefore, this graph can be used for being processed by a \ac{GNN} as done in~\cite{DBLP:conf/corl/TolstayaGPP0R19,tolstaya2020learning,DBLP:conf/icra/GosrichMLPYR022}.
\subsection{Motivation}
The definition of distributed controllers for intelligent systems can be divided into two major approaches: 
 \emph{manual design}, where the application is conceived and described through a programming language, 
 and \emph{automatic design}~\cite{brambilla2013swarm}, which employs machine learning techniques to synthesise distributed controllers. 
%
Our work lies between these two approaches, specifically in the areas of field-based coordination and \ac{MAARL} with the use of \acp{GNN}.

@@

Let $\Gamma$ be the aggregate program that takes a graph $G_t$, representing the participating nodes and their neighbourhood relations at time $t$, as input. 

@@

The feature vector $o_v$ of which is composed the graph $G'_t$ used by the policy $\pi_{gnn}$ is described by:
 $o_v = (\theta^v_t, \omega)$. \lukas{we used $o_v$ to describe the observation of node $v$ in section II.C - so this would not include field information, right?}

@@

 where $\omega$ are other local information gather from sensors (like temperature, distances from neighbourhood, etc.). 

>>>>

On these minimal operators, 
 it is then possible to build \emph{self-organizing} coordination blocks.
 One of the founding blocks is the \emph{gradient} essential for 
 \emph{information} flows~\cite{DBLP:conf/saso/WolfH07}.
%
This operator generates a numerical field that represents the minimum distance from a source zone. 
%
In other words, it maps a Boolean field (where ``true'' indicates the presence of a source node and ``false'' indicates its absence) to a distance field that indicates the proximity to the nearest source. In ScaFi (a scala implementation of aggregate computing) it is simply expressed as:
\begin{lstlisting}
def gradient(source: Boolean): Double
\end{lstlisting}
Along the gradient, it is possible to cast information within the system, accumulating value during the programming. This pattern is called \emph{gradient-cast} (or \emph{G}) 
\begin{lstlisting}
def G[V](source: Boolean, value: V, acc: V => V): V
\end{lstlisting}
\ga{more description here}
this is used in the use case to accumulate the distance from the phenomena that is tracked
\lukas{still not sure how the coordination works or if it has been explained in the above paragraph.}
\subsection{Graph Neural Networks}
%Over the years, several different neural networks have been proposed to solve specific tasks, ranging from 
%  simple \ac{MLP}~\cite{werbos1975beyond} to \ac{CNN}~\cite{oshea2015introduction}.
%  \lukas{more examples? }
%%
%Whereas \ac{CNN} is designed to be used with image and spatial-like data (e.g., audio), 
 \ac{GNN} is a novel neural network model used to process graph-structured data with deep learning approaches.
%
Let $G = (V, E)$ a graph  
where $E$ defines the neighbourhood relations for each participating node, 
and $V$ identifies the nodes present in the graph. %at that time.  \lukas{no time in the definition}
% 
Each node $v \in V$ is associated with an observation (or feature set) $o_v$.
%\lukas{Over time this graph $G$ can change and therefore we define $G_t$ as the graph $G$ at a specific time $t$.} 
%
Given $o_v$, %a feature vector associated with each node $v \in V$, 
 the goal of a \ac{GNN} is to learn the node embedding $h_v$ for each node $v \in V$. The node embedding $h_v$ describes the node in the network and summarises the geometric properties of the graph in this location, allowing for comparison of various nodes in the graph.
 In modern \acp{GNN}, the node embedding $h_v$ is computed by aggregating information from the node's neighbours $\mathcal{N}_G(v)$,
  and then combining it with the node's current embedding $h_v$.
 The graph is further partitioned into layers with the node $v$ at its bottom layer $0$. With each traversal of an edge $e \in E$ from $v$, the number of layers increased
Formally, a \ac{GNN} can be defined by two phases:
\begin{equation}
a_{v}^{(k)}=\operatorname{AGGREGATE}^{(k)}\left(\left\{h_{u}^{(k-1)}: u \in \mathcal{N}_G(v)\right\}\right)   
\end{equation}

\begin{equation}
h_{v}^{(k)}=\operatorname{COMBINE}^{(k)}\left(h_{v}^{(k-1)}, a_{v}^{(k)}\right)
\end{equation}

where $h_{v}^{k}$ is the embedding of node $v$ within the $k$-th layer, 
 $\mathcal{N}_G(v)$ is the set of neighbours of node $v$ computed from $E$.
%
$h_v{v}^0$ is the initial embedding of node $v$ and it is usually set to the node's feature vector $f_v$.
%
The differential part comes into play in the $\operatorname{COMBINE}$ function, 
 which is usually a differentiable function such as a neural network layer.
%
$\operatorname{AGGREGATE} $ instead is a function that aggregates the information from the neighbours of a node $v$
  and it could be a simple sum, max or sum of products. More complex aggregation functions are available~\cite{pellegrini2020learning}.
%
This formulation allows GNNs to effectively process and extract features from graph-structured data by iteratively aggregating and transforming information from the node's neighbours.
\lukas{do we need definitions of $o_v$, $h_v^*$ over time $t$?}

@@

This model can be used then to express the evolution of the system in time.
Specifically, starting from a global state $\mathcal{S}^P_t$, the next state $\mathcal{S}^P_{t+1}$ is defined as:
\begin{equation}
\mathcal{A}^P_t = \pi(\xi(S^P_t))
\end{equation}
\begin{equation}
\mathcal{S}^P_{t+1} = \mathcal{T}(\mathcal{S}^P_t, \mathcal{A}^P_t)
\end{equation}
Given a time $t$, the system can be also represented as a graph $G_t = (V_t, E_t)$, 
  where $E_t$ is built from $\mathcal{N}$.
  Each node is then decorated with the local observation $f_v = o_v$. 
  Therefore, this graph can be used for being processed by a \ac{GNN} as done in~\cite{DBLP:conf/corl/TolstayaGPP0R19,tolstaya2020learning,DBLP:conf/icra/GosrichMLPYR022}.
\subsection{Motivation}
The definition of distributed controllers for intelligent systems can be divided into two major approaches: 
 \emph{manual design}, where the application is conceived and described through a programming language, 
 and \emph{automatic design}~\cite{brambilla2013swarm}, which employs machine learning techniques to synthesise distributed controllers. 
%
Our work lies between these two approaches, specifically in the areas of field-based coordination and \ac{MAARL} with the use of \acp{GNN}.

@@

Let $\Gamma$ be the aggregate program that takes a graph $G_t$ decorated by $o_v$ (i.e., the local observation)), 
 representing the participating nodes and their neighbourhood relations at time $t$, as input. 

@@

From this field, we construct the feature vector $f_v$ for each node $v$ in $G_t$ as follows:
 $f_v = (\theta^v_t, o_t)$. \lukas{we used $o_v$ to describe the observation of node $v$ in section II.C - so this would not include field information, right?}

@@

 %where $\omega$ are other local information gather from sensors (like temperature, distances from neighbourhood, etc.). 
